---
title: "Lab Assignment 3: Parameter Estimation and Unbiasedness of Estimators"
author: "Team [Team Number]"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Work Breakdown Structure

- Problem 1 - Lidiia Sokha
- Problem 2 - Nazar Pasichnyk
- Problem 3 - Viktoriia Kukurudza

**Total:** 100%

# Introduction

```{r seed}
# Set seed for reproducibility
set.seed(39)
```

---

# Part I: Parameter Estimation

## Problem 1: Exponential Distribution Confidence Intervals

### Problem Formulation and Discussion

**Objective:**

**Parameters:**
- θ = team_id/10
- α levels: 0.1, 0.05, 0.01
- Sample sizes: n (to be varied)
- Number of repetitions: m (to be varied)

**Discussion:**

### Method 1: Exact Distribution Using Chi-Square

#### Theoretical Background

#### R Implementation

```{r problem1-method1}
# Implementation for Method 1
```

#### Results and Statistics

```{r problem1-method1-results}
# Calculate statistics and create visualizations
```

### Method 2: Normal Approximation with Known Variance

#### Theoretical Background

#### R Implementation

```{r problem1-method2}
# Implementation for Method 2
```

#### Results and Statistics

```{r problem1-method2-results}
# Calculate statistics and create visualizations
```

### Method 3: Normal Approximation with Solved Inequality

#### Theoretical Background

#### R Implementation

```{r problem1-method3}
# Implementation for Method 3
```

#### Results and Statistics

```{r problem1-method3-results}
# Calculate statistics and create visualizations
```

### Method 4: Student t-Distribution Approach

#### Theoretical Background

#### R Implementation

```{r problem1-method4}
# Implementation for Method 4
```

#### Results and Statistics

```{r problem1-method4-results}
# Calculate statistics and create visualizations
```

### Task (a): Verification of Coverage Probability

```{r problem1-verification}
# Verify that confidence intervals contain θ approximately 100(1-α)% of times
```

#### Coverage Probability Results

#### Visualization

```{r problem1-coverage-plots}
# Histograms and plots showing coverage
```

### Task (b): Comparison of Precision (Interval Lengths)

```{r problem1-precision}
# Compare lengths of confidence intervals
```

#### Precision Comparison Results

#### Visualization

```{r problem1-precision-plots}
# Plots comparing interval lengths
```

### Task (c): Recommendation and Justification

**Best Method:**

**Justification:**

**Conclusion:**

---

## Problem 2: Poisson Distribution Confidence Intervals

### Problem Formulation and Discussion

**Objective:**

**Parameters:**
- θ = team_id/10
- α levels: 0.1, 0.05, 0.01
- Sample sizes: n (to be varied)
- Number of repetitions: m (to be varied)

**Discussion:**

### Method 2: Normal Approximation with Known Variance

#### Theoretical Background

#### R Implementation

```{r problem2-method2}
# Implementation for Method 2 (Poisson)
```

#### Results and Statistics

```{r problem2-method2-results}
# Calculate statistics and create visualizations
```

### Method 3: Normal Approximation with Solved Inequality

#### Theoretical Background

#### R Implementation

```{r problem2-method3}
# Implementation for Method 3 (Poisson)
```

#### Results and Statistics

```{r problem2-method3-results}
# Calculate statistics and create visualizations
```

### Method 4: Student t-Distribution Approach

#### Theoretical Background

#### R Implementation

```{r problem2-method4}
# Implementation for Method 4 (Poisson)
```

#### Results and Statistics

```{r problem2-method4-results}
# Calculate statistics and create visualizations
```

### Verification of Coverage Probability

```{r problem2-verification}
# Verify that confidence intervals contain θ approximately 100(1-α)% of times
```

#### Coverage Probability Results

#### Visualization

```{r problem2-coverage-plots}
# Histograms and plots showing coverage
```

### Comparison of Precision (Interval Lengths)

```{r problem2-precision}
# Compare lengths of confidence intervals
```

#### Precision Comparison Results

#### Visualization

```{r problem2-precision-plots}
# Plots comparing interval lengths
```

### Recommendation and Justification

**Best Method:**

**Justification:**

**Conclusion:**

---

# Part II: Unbiasedness of Estimators

## Problem 3: Analysis of Sample Variance Estimators

### Problem Formulation and Discussion

**Objective:**

**Variance Estimators:**

$$\sigma^2_n = \frac{1}{n}\sum_{i=1}^{n}(X_i - \bar{X})^2$$

$$\sigma^2_{n-1} = \frac{1}{n-1}\sum_{i=1}^{n}(X_i - \bar{X})^2$$

**Discussion:**

### Task (a): Code to Calculate Variance

```{r problem3-variance-code}
# Function to calculate both variance estimators
```

### Task (b): Calculate Estimators for Different Sample Sizes

#### n = 10

```{r problem3-n10}
# Calculate σ²_n and σ²_{n-1} for n=10
```

#### n = 50

```{r problem3-n50}
# Calculate σ²_n and σ²_{n-1} for n=50
```

#### n = 100

```{r problem3-n100}
# Calculate σ²_n and σ²_{n-1} for n=100
```

#### n = 1000

```{r problem3-n1000}
# Calculate σ²_n and σ²_{n-1} for n=1000
```

#### Summary Table

```{r problem3-summary-table}
# Create summary table of results
```

### Task (c): Calculate Biases

```{r problem3-biases}
# Calculate Bias(σ²_n) = E(σ²_n) - σ²
# Calculate Bias(σ²_{n-1}) = E(σ²_{n-1}) - σ²
```

#### Bias Results

#### Visualization

```{r problem3-bias-plots}
# Plot biases for different sample sizes
```

### Task (d): Commentary on Results for Different Values of n

**Observations:**

**Analysis:**

### Task (e): Analytical Derivation of Expected Values

#### Expected Value of σ²_n

**Derivation:**

$$E(\sigma^2_n) = E\left[\frac{1}{n}\sum_{i=1}^{n}(X_i - \bar{X})^2\right]$$

**Result:**

#### Expected Value of σ²_{n-1}

**Derivation:**

$$E(\sigma^2_{n-1}) = E\left[\frac{1}{n-1}\sum_{i=1}^{n}(X_i - \bar{X})^2\right]$$

**Result:**

### Task (f): Mathematical Proof of Unbiasedness

#### Analysis of σ²_n

**Bias:**

**Conclusion:**

#### Analysis of σ²_{n-1}

**Bias:**

**Conclusion:**

### Task (g): Commentary on Theoretical vs Practical Results

**Theoretical Findings:**

**Practical Findings:**

**Comparison:**

**Conclusion:**

---

# Overall Conclusions

## Summary of Findings

### Part I: Parameter Estimation

### Part II: Unbiasedness of Estimators

## Final Remarks

---

